---
title: "Fecundity analyses"
author: "Zajitschek et al."
date: "2023-06-01"
output: 
  workflowr::wflow_html:
   code_folding: hide
editor_options: 
  chunk_output_type: console
---

## Load libraries
```{r load-libraries, message=FALSE, warning=FALSE, results="hide"}
library(readr)      # read data
library(dplyr)      # data structure
library(reshape2)   # data structure
library(rcompanion) # mean and median (+- bootstrapped CI)
library(ggplot2)    # plot
library(sjPlot)     # model plots & diagnostics
library(ggeffects)  # plot
library(car)        # anova-style results table for glm
library(bbmle)      # IC tab
library(robustbase) # robust glm
library(splines)    # spline functions
library(MuMIn)      # model selection
library(lmtest)     # for F1 model formulation
library(sandwich)   # for F1 model formulation
library(multcomp)   # for F1 multiple comparisons

```

## Load fecundity data (parental generation F0, and F1 offspring)
```{r import-fecundity, message=FALSE}
# F0 fecundity 
fec.F0 <- read_csv("./data/fecF0.csv")
colnames(fec.F0)[1] <- "generation"
fec.F0$diet_num <- as.numeric(as.character(fec.F0$diet))
fec.F0$diet <- as.factor(fec.F0$diet)
fec.F0$vial_id <- as.factor(fec.F0$vial_id)
fec.F0$rep <- as.factor(fec.F0$rep)

# F1 fecundity 
fec.F1 <- read_csv("./data/fecF1.csv")
colnames(fec.F1)[1] <- "generation"
fec.F1$par_diet <- as.factor(fec.F1$par_diet)  #'par_diet' stands for 'parental diet'
fec.F1$vial_id <- as.factor(fec.F1$vial_id)
fec.F1$rep <- as.factor(fec.F1$rep)
```
## 1) F0  FECUNDITY

## Explore F0 fecundity data
```{r explore-fecundity, message=FALSE, warning=FALSE}
#plot(fec.F0$eggs) # ok: no off-values/outliers
#max(fec.F0$eggs)  # max. eggs = 141

print(paste0("5 measurement ages [days]: ", list(unique(fec.F0$age))))
 # 5 measurement ages: 5, 12, 19, 26, 33 days old

#Summary statistics and dataset for further analyses, containing sums of counted eggs per vial
fec.F0.a <- fec.F0 %>% 
  group_by(temperature, diet_num, vial_id) %>% 
  summarise(., vial_sum = sum(eggs), n_ages = n_distinct(age)) 
fec.F0.a %>% group_by(temperature, diet_num) %>%
  summarise(., mean_sum_fec = mean(vial_sum), se = sd(vial_sum)/sqrt(10), sd = sd(vial_sum), 
   median = median(vial_sum), n_vials = n_distinct(vial_id))
```

## Calculate mean fecundity (across vials) with bootstrapped CI
```{r mean-fecundity}
boot.fec.F0 <- groupwiseMean(vial_sum ~ temperature*diet_num, data = fec.F0.a, boot = T, bca = T)  #5000 bootstrap replicates for CI
boot.fec.F0
```

## Plot F0 mean fecundity (not in manuscript)
```{r plot-F0-mean-fecundity}
mycols <- colors()[c(30, 62, 119, 504, 554)]

fec.F0.plot <- ggplot(data= boot.fec.F0, aes(x= as.factor(diet_num), y= Boot.mean, fill= as.factor(temperature))) +  
	#scale_fill_grey(start=0.8, end=0.3, name = "Temperature") +
	scale_fill_manual(values= mycols, name = "Temperature") +
     	geom_bar(position=position_dodge(), stat = "identity", colour="black") + 	
	geom_errorbar(aes(ymin= Bca.lower, ymax= Bca.upper), width=.4, , colour="black", position=position_dodge(0.9)) +
	ylab("Fecundity (mean # of eggs)")  + 
	xlab("Diet")  + 
	ggtitle("F0 Fecundity\n(vial means over first 5 weeks \U00B1 95% CI)") + 
	theme_bw(base_size = 18)+
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(), panel.border = element_blank(), 
	axis.line.x = element_line(), axis.line.y = element_line()) 
fec.F0.plot 
```

## F0 FECUNDITY ANALYSES

### Model specification, results (Table S8), and comparison

```{r fecundity-totalsum-analysis}
fec.F0.glm.num.1 <- glm(vial_sum ~ bs(temperature)*diet_num +
                          bs(temperature)*I(diet_num^2), data= fec.F0.a) 
#plot(fec.F0.glm.num.1) #problematic QQ-plot for upper quantiles

#Square-root transformed response 
#Refit above with new response variable 'vial_sum2'
fec.F0.a$vial_sum2 <- (fec.F0.a$vial_sum)^(1/2)

fec.F0.glm.num.2 <- glm(vial_sum2 ~ bs(temperature)*diet_num +
                          bs(temperature)*I(diet_num^2), data= fec.F0.a) 
#plot(fec.F0.glm.num.2) #better QQ-plot


#Robust GLM
fec.F0.glm.4 <- glmrob(vial_sum2 ~ bs(temperature)*diet_num +
                          bs(temperature)*I(diet_num^2), family = gaussian(), data = fec.F0.a)
summary(fec.F0.glm.4)

#Remove highest interaction term from initial model
fec.F0.glm.4b <- update(fec.F0.glm.4, . ~ . - bs(temperature):I(diet_num^2))

#Compare models
anova(fec.F0.glm.4b, fec.F0.glm.4, test = "Wald")

```

### Plot Figure 2: F0 female fecundity

```{r Figure-2}

predrob.4 <- ggpredict(fec.F0.glm.4, c("temperature", "diet_num")) 
predFecF0_plot <- ggplot(predrob.4, aes(x, predicted))  +
  geom_point(data = fec.F0.a, aes(x = temperature, y = vial_sum2, shape  = factor(diet_num), colour = factor(diet_num), stroke=NA, alpha = 0.1), size = 3, position= position_jitterdodge(dodge.width = 0.8, jitter.width = 0.1), inherit.aes = F) + 
    geom_line(aes(linetype= group, color= group), linewidth = 1.5) +
    geom_ribbon(aes(ymin=conf.low, ymax=conf.high, fill=group), alpha=0.15) +
    scale_color_manual(name  ="Diet", values=c("#F8766D", "#619CFF", "#00BA38"), labels=c("restricted", "standard", "rich")) +
  scale_fill_manual(name  ="Diet", values=c("#F8766D", "#619CFF", "#00BA38"), labels=c("restricted", "standard", "rich"))  +
  scale_shape_discrete(name  ="Diet", labels=c("restricted", "standard", "rich")) +
    scale_linetype_manual(name  ="Diet", values = c("solid", "dashed", "dotted"), labels=c("restricted", "standard", "rich")) +
  scale_x_continuous(breaks=c(21, 23, 25, 27, 29)) + 
  labs(title = "", color = "Diet", y = "Total fecundity\n(square-root, weeks 1 to 5)", x = "Temperature [degree Celsius]") + theme_classic(base_size = 23) + 
 guides(alpha = "none")

predFecF0_plot
```


## 2) F1 FECUNDITY

### Explore F1 fecundity data

```{r explore-F1-fecundity, message=FALSE, warning=FALSE}
#plot(fec.F1$eggs) # ok: no off-values/outliers
#max(fec.F1$eggs)  # max. eggs = 217

print(paste0("5 measurement ages [days]: ", list(unique(fec.F1$age))))
 # 5 measurement ages: 4, 11, 18, 25, 32 days old

fec.F1.a <- fec.F1 %>% 
  group_by(par_temp, par_diet, par_eff, vial_id) %>% 
  summarise(., vial_sum = sum(eggs), n_ages = n_distinct(age)) 
fec.F1.a %>% group_by(par_temp, par_diet, par_eff) %>%
  summarise(., mean_sum_fec = mean(vial_sum), se = sd(vial_sum)/sqrt(length(vial_sum)), 
  sd = sd(vial_sum), median = median(vial_sum), n_vials = n_distinct(vial_id))
```

### Calculate F1 mean fecundity (across vials) with bootstrapped CI

```{r F1-mean-fecundity}
F1.fec.CI <- groupwiseMean(vial_sum ~ par_temp*par_diet*par_eff, data = fec.F1.a, boot = F, bca = F)  
F1.fec.CI$Trad.lower[F1.fec.CI$Trad.lower < 0] <- 0  #set negative lower CI range values to zero
```

### Plot F1 mean fecundity (not in manuscript)

```{r plot-F1-mean-fecundity}
mycols <- colors()[c(30, 62, 119, 504, 554)]

fec.F1.plot <- ggplot(data= F1.fec.CI, aes(x= as.factor(par_diet), y= Mean, fill= as.factor(par_temp))) +  
	#scale_fill_grey(start=0.8, end=0.3, name = "Temperature") +
	scale_fill_manual(values= mycols, name = "Parental\ntemperature") +
     	geom_bar(position=position_dodge(), stat = "identity", colour="black") + 	
	geom_errorbar(aes(ymin= Trad.lower, ymax= Trad.upper), width=.4, , colour="black", position=position_dodge(0.9)) + 
	ylab("Fecundity (mean # of eggs)")  + 
	xlab("Parental diet")  + 
	ggtitle("F1 Fecundity\n(vial means over first 5 weeks \U00B1 95% CI)") + 
	theme_bw(base_size = 18)+
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(), panel.border = element_blank(), 
	axis.line.x = element_line(), axis.line.y = element_line()) +
  facet_grid(vars(par_eff), vars(par_temp))

fec.F1.plot 
```

## F1 FECUNDITY ANALYSES

### Additional data preparation

```{r F1-fecundity-totalsum-analysis-prepglm}
fec.F1.a$par_diet_num <- as.numeric(as.character(fec.F1.a$par_diet))
fec.F1.a$vial_sum2 <- (fec.F1.a$vial_sum)^(1/2)
```

### Robust estimation of effects on F1 fecundity (Table S9)

```{r F1-r F1-fecundity-totalsum-analysis-glmrob, eval=FALSE, include=FALSE}
#Robust glms
fec.F1.glm0 <- glmrob(vial_sum ~ par_eff*bs(par_temp)*par_diet_num +
               par_eff*bs(par_temp)*I(par_diet_num^2), family = gaussian(), data = fec.F1.a) #fec.F1.glm0 doesn't converge

# We use square-root of fecundity as the response variable to get the model to converge
fec.F1.glm <- glmrob(vial_sum2 ~ par_eff*bs(par_temp)*par_diet_num +
              par_eff*bs(par_temp)*I(par_diet_num^2), family = gaussian(), data = fec.F1.a)
summary(fec.F1.glm) 

```

```{r F1-fecundity-totalsum-analysis-aov}

#All independent variables as factor to make the model less complex; separate for each level of par_eff
fec.F1.a$par_temp_fac <- as.factor(fec.F1.a$par_temp)

# Maternal effects with robust covariance matrix 
fec.F1.a.mat <- subset(fec.F1.a, par_eff=="mat")

fec.F1.glm.rev1c.mat <- aov(vial_sum ~ par_temp_fac*par_diet, data = fec.F1.a.mat)
fec.F1.glm.rev1c.mat.coef <- coeftest(fec.F1.glm.rev1c.mat, vcov = vcovHC)

fec.F1.glm.rev1c.mat2 <- aov(vial_sum ~ par_temp_fac+par_diet, data = fec.F1.a.mat)

# Model comparison
waldtest(fec.F1.glm.rev1c.mat, fec.F1.glm.rev1c.mat2, vcov = vcovHC, test = "F")

Tukey. <- contrMat(table(fec.F1.a.mat$par_temp_fac), "Tukey")
K1. <- cbind(Tukey., matrix(0, nrow = nrow(Tukey.), ncol = ncol(Tukey.)), matrix(0, nrow = nrow(Tukey.), ncol = ncol(Tukey.)))

rownames(K1.) <- paste(levels(fec.F1.a$par_diet)[1], rownames(K1.), sep = ":")

K2. <- cbind(matrix(0, nrow = nrow(Tukey.), ncol = ncol(Tukey.)), Tukey., matrix(0, nrow = nrow(Tukey.), ncol = ncol(Tukey.)))
rownames(K2.) <- paste(levels(fec.F1.a$par_diet)[2], rownames(K2.), sep = ":")

K3. <- cbind(matrix(0, nrow = nrow(Tukey.), ncol = ncol(Tukey.)), matrix(0, nrow = nrow(Tukey.), ncol = ncol(Tukey.)),Tukey.)
rownames(K3.) <- paste(levels(fec.F1.a$par_diet)[3], rownames(K3.), sep = ":")
K. <- rbind(K1., K2., K3.)
colnames(K.) <- c(colnames(Tukey.), colnames(Tukey.), colnames(Tukey.))

#todo: summary(glht(fec.F1.glm.rev1c.mat, linfct = K. %*% X., vcov = sandwich::vcovHC(fec.F1.glm.rev1c.mat)), test = adjusted("BH"))  
 

# Paternal effects with robust covariance matrix 
fec.F1.a.pat <- subset(fec.F1.a, par_eff=="pat")

fec.F1.glm.rev1c.pat <- aov(vial_sum ~ par_temp_fac*par_diet, data = fec.F1.a.pat)
coeftest(fec.F1.glm.rev1c.pat, vcov = sandwich::vcovHC)

fec.F1.glm.rev1c.pat2 <- aov(vial_sum ~ par_temp_fac+par_diet, data = fec.F1.a.pat)

# Model comparison
waldtest(fec.F1.glm.rev1c.pat, fec.F1.glm.rev1c.pat2, vcov = vcovHC, test = "F")

Tukey <- contrMat(table(fec.F1.a.pat$par_temp_fac), "Tukey")
K1 <- cbind(Tukey, matrix(0, nrow = nrow(Tukey), ncol = ncol(Tukey)), matrix(0, nrow = nrow(Tukey), ncol = ncol(Tukey)))

rownames(K1) <- paste(levels(fec.F1.a$par_diet)[1], rownames(K1), sep = ":")

K2 <- cbind(matrix(0, nrow = nrow(Tukey), ncol = ncol(Tukey)), Tukey, matrix(0, nrow = nrow(Tukey), ncol = ncol(Tukey)))
rownames(K2) <- paste(levels(fec.F1.a$par_diet)[2], rownames(K2), sep = ":")

K3 <- cbind(matrix(0, nrow = nrow(Tukey), ncol = ncol(Tukey)), matrix(0, nrow = nrow(Tukey), ncol = ncol(Tukey)),Tukey)
rownames(K3) <- paste(levels(fec.F1.a$par_diet)[3], rownames(K3), sep = ":")
K <- rbind(K1, K2, K3)
colnames(K) <- c(colnames(Tukey), colnames(Tukey), colnames(Tukey))
 
#todo: summary(glht(fec.F1.glm.rev1c.pat, linfct = K %*% X, vcov = sandwich::vcovHC(fec.F1.glm.rev1c.pat)), test = adjusted("BH")) 

```

### Plot Figures 3 and 4

```{r F1-fecundity-model-predictions-plot-rev}

#Maternal

predrob.F1.mat <- ggpredict(fec.F1.glm.rev1c.mat, c("par_temp_fac", "par_diet"), vcov.fun = "vcovHC", vcov.type = "HC")

predFecF1.mat <- plot(predrob.F1.mat, rawdata = T, facet = T, connect.lines = T, ci.style = "errorbar", line.size = 1.5, dot.size = 5)  

predFecF1_plot.mat <- predFecF1.mat  + 
  labs(title = "Maternal diet", y = "Total fecundity\n(weeks 1 to 5)", x = "Temperature [degree Celsius]", color = "Parental\ndiet") + theme_classic(base_size = 23) 
predFecF1_plot.mat

#Paternal

predrob.F1.pat <- ggpredict(fec.F1.glm.rev1c.pat, c("par_temp_fac", "par_diet"), vcov.fun = "vcovHC", vcov.type = "HC")
plot(predrob.F1.pat, rawdata = T, facet = T, connect.lines = T, ci.style = "errorbar", line.size = 1.5, dot.size = 5)-> predFecF1.pat
predFecF1_plot.pat <- predFecF1.pat + 
  labs(title = "Paternal diet", y = "Total fecundity\n(weeks 1 to 5)", x = "Temperature [degree Celsius]", color = "Parental\ndiet") + theme_classic(base_size = 23)
predFecF1_plot.pat
```
