---
title: "Mating behaviour analyses"
author: "Zajitschek et al."
date: "2023-06-01"
output: 
  workflowr::wflow_html:
   code_folding: hide
editor_options: 
  chunk_output_type: console
---

## Load libraries
```{r load-libraries, message=FALSE, warning=FALSE, results="hide"}
library(readr)      # read data
library(dplyr)      # data structure
library(reshape2)   # data structure
library(rcompanion) # mean and median (+- bootstrapped CI)
library(ggplot2)    # plot
library(sjPlot)     # model plots & diagnostics
library(grid)       # figure arrangement
library(gtable)     # figure arrangement
library(gridExtra)  # figure arrangement
library(lme4)       # linear mixed models (LMM)
library(lmerTest)   # significance testing for LMM
library(multcomp)   # multiple comparison correction
library(lsmeans)
library(pbkrtest)
library(predictmeans) #diagnostic plots for (g)lmer: residplot
library(gamm4)      # additive mixed models
library(robustbase)
library(effects)
library(MuMIn)
library(gratia)     # diagnostics of gam models
library(DHARMa)
library(splines)

```

## Load male mating behaviour data (parental generation F0, F1, F2, and F3)

```{r import-matingbehaviour, message=FALSE}

# F0 male mating behaviour 
mb.f0 <- read_csv("./data/malebehavF0.csv")
mb.f0a <- mb.f0[1:210, ]
mb.f0a$treat <- interaction(mb.f0a$diet, mb.f0a$temp)
mb.f0a$tray <- interaction(mb.f0a$treat, ifelse(mb.f0a$male_id < 8, c("a"), c("b")) )
mb.f0a$male_id <- factor(mb.f0a$male_id)
mb.f0a$diet_fac <- factor(mb.f0a$diet)
mb.f0a$temp_fac <- factor(mb.f0a$temp)
mb.f0a$replicate_vial_sep <- interaction(mb.f0a$treat, mb.f0a$replicate_vial)
mb.f0a$indobs <- factor(1:nrow(mb.f0a))

# F1 male mating behaviour 
mb.f1 <- read_csv("./data/malebehavF1.csv")
mb.f1a <- mb.f1
mb.f1a$treat <- interaction(mb.f1a$parental_effects, mb.f1a$parental_diet, mb.f1a$parental_temp)
mb.f1a$tray <- interaction(mb.f1a$treat, ifelse(mb.f1a$male_id < 8, c("a"), c("b")) )
mb.f1a$male_id <- factor(mb.f1a$male_id)
mb.f1a$parental_diet_fac <- factor(mb.f1a$parental_diet)
mb.f1a$parental_temp_fac <- factor(mb.f1a$parental_temp)
mb.f1a$replicate_vial_sep <- interaction(mb.f1a$treat, mb.f1a$replicate_vial)
mb.f1a$indobs <- factor(1:nrow(mb.f1a))
mb.f1.analysis <- mb.f1a %>% 
  filter(!is.na(latency_to_mate))

# F2 male mating behaviour 
mb.f2 <- read.csv("./data/malebehavF2.csv")  
mb.f2a <- mb.f2
mb.f2a$treat <- interaction(mb.f2a$parental_effects, mb.f2a$parental_diet, mb.f2a$parental_temp)
mb.f2a$tray <- interaction(mb.f2a$treat, ifelse(mb.f2a$male_id < 8, c("a"), c("b")) )
mb.f2a$male_id <- factor(mb.f2a$male_id)
mb.f2a$parental_diet_fac <- factor(mb.f2a$parental_diet)
mb.f2a$parental_temp_fac <- factor(mb.f2a$parental_temp)
mb.f2a$replicate_vial_sep <- interaction(mb.f2a$treat, mb.f2a$replicate_vial)
mb.f2a$indobs <- factor(1:nrow(mb.f2a))

# F3 male mating behaviour 
mb.f3 <- read.csv("./data/malebehavF3.csv")  
mb.f3a <- mb.f3
mb.f3a$treat <- interaction(mb.f3a$parental_effects, mb.f3a$parental_diet, mb.f3a$parental_temp)
mb.f3a$tray <- interaction(mb.f3a$treat, ifelse(mb.f3a$male_id < 8, c("a"), c("b")) )
mb.f3a$male_id <- factor(mb.f3a$male_id)
mb.f3a$parental_diet_fac <- factor(mb.f3a$parental_diet)
mb.f3a$parental_temp_fac <- factor(mb.f3a$parental_temp)
mb.f3a$replicate_vial_sep <- interaction(mb.f3a$treat, mb.f3a$replicate_vial)
mb.f3a$indobs <- factor(1:nrow(mb.f3a))

```

## Data for latency to mate and mating duration in F0

```{r malelatency-F0, message=FALSE}

mb.f0 <- mb.f0a %>% 
  filter(!is.na(latency_to_mate)) %>%
  group_by(temp_fac, diet_fac) %>%   
  summarise(mean_latency_to_mate = mean(latency_to_mate, na.rm = TRUE),  
            sd_latency_to_mate = sd(latency_to_mate), 
            n_latency_to_mate = n(),  
            SE_latency_to_mate = sd(latency_to_mate)/sqrt(n()),
		mean_mating_duration = mean(mating_duration, na.rm = TRUE),  
            sd_mating_duration = sd(mating_duration), 
            n_mating_duration = n(),  
            SE_mating_duration = sd(mating_duration)/sqrt(n())) 

print(mb.f0, n= 30)

```

## Data for F0 mating success

```{r malematingsuccessF0-data}

mb.f0.yn <- mb.f0a %>% 
  group_by(temp_fac, diet_fac) %>%   
  summarise(mating_success = sum(mated == "yes") / 14) 
print(mb.f0.yn, n= 60)

```

## MALE MATING BEHAVIOUR ANALYSES

## F0 MALE MATING BEHAVIOUR

### F0 Latency to mate, GAMM: model comparison (table S10) and results (inspect code for model specifications)

```{r latency-F0-gamm}

mb.f0.analysis <- mb.f0a %>% 
  filter(!is.na(latency_to_mate))

# Full model: residuals are right-skewed
matlatF0.gamm0 <- gamm4(latency_to_mate ~ s(temp, k= 5, by= diet_fac, bs="cr") + diet_fac, random= ~ (1|replicate_vial_sep), data= mb.f0.analysis )
# Diagnostic plot
appraise(matlatF0.gamm0$gam)

# Better with log-transformed response
matlatF0.gamm1 <- gamm4(log(latency_to_mate) ~ s(temp, k= 5, by= diet_fac, bs="cr") + diet_fac, random= ~ (1|replicate_vial_sep), data= mb.f0.analysis )
# Diagnostic plot
appraise(matlatF0.gamm1$gam)

# Model without linear diet term
matlatF0.gamm1b <- gamm4(log(latency_to_mate) ~ s(temp, k= 5, by= diet_fac, bs="cr"), random= ~ (1|replicate_vial_sep), data= mb.f0.analysis )

# Model without diet-specific smooth terms
matlatF0.gamm2 <- gamm4(log(latency_to_mate) ~s(temp, k= 5, bs="cr") + diet_fac, random= ~ (1|replicate_vial_sep), data= mb.f0.analysis)

# Model without any diet effects
matlatF0.gamm3 <- gamm4(log(latency_to_mate) ~s(temp, k= 5, bs="cr"), 
			    random= ~ (1|replicate_vial_sep), data= mb.f0.analysis)

# Model with linear temperature term only
matlatF0.gamm3b <- gamm4(log(latency_to_mate) ~ temp, 
			    random= ~ (1|replicate_vial_sep), data= mb.f0.analysis)

# Model with linear temperature and diet terms only
matlatF0.gamm3c <- gamm4(log(latency_to_mate) ~ temp*diet, 
			    random= ~ (1|replicate_vial_sep), data= mb.f0.analysis)

matlatF0.gamm.null <- gamm4(log(latency_to_mate) ~ 1, 
			    random= ~ (1|replicate_vial_sep), data= mb.f0.analysis )

# AIC model table (Table S10)
model.sel(matlatF0.gamm1$mer, matlatF0.gamm1b$mer, matlatF0.gamm2$mer, matlatF0.gamm3$mer, matlatF0.gamm3b$mer, matlatF0.gamm3c$mer, matlatF0.gamm.null$mer, rank = AIC)

summary(matlatF0.gamm3$gam)

```

### Plot Figure 5 

```{r Figure-5}
# Fitted.values(matlatF0.gamm3$gam)
x_latF0dat <- data.frame(temp = seq(21, 29, length.out = 100))
pred_latF0dat <- predict(matlatF0.gamm3$gam, x_latF0dat, se.fit = TRUE)

new_latF0dat <- cbind(x_latF0dat, pred_latF0dat)

# Calculate CIs 
new_latF0dat <- within(new_latF0dat, {
    lower = fit - 1.96*se.fit
    upper = fit + 1.96*se.fit
    fit_orig = exp(fit)
    lower_orig = exp(lower)
    upper_orig = exp(upper)
})

latF0_gamm3_plot <- ggplot(new_latF0dat, aes(x = temp, y = fit_orig)) + 
  geom_point(data = mb.f0.analysis, aes(x = temp, y = latency_to_mate, color = factor(diet), shape  = factor(diet)), size = 3.2, alpha = 0.6, position= position_jitterdodge(dodge.width = 0.8, jitter.width = 0.1)) + 
  geom_ribbon(aes(ymin = lower_orig, ymax = upper_orig), alpha=0.1) +
  geom_line(size = 1.1) +
  scale_x_continuous(breaks=c(21, 23, 25, 27, 29)) + 
  scale_color_manual(name  ="Diet", values=c("#F8766D", "#619CFF", "#00BA38"), labels=c("restricted", "standard", "rich")) +
  scale_fill_manual(name  ="Diet", values=c("#F8766D", "#619CFF", "#00BA38"), labels=c("restricted", "standard", "rich"))  +
  scale_shape_discrete(name  ="Diet", labels=c("restricted", "standard", "rich")) +
  labs(title = "", color = "Diet", y = "Latency to mate [min]", x = "Temperature [degree Celsius]") + theme_classic(base_size = 23) 

latF0_gamm3_plot

```

### F0 Mating duration, GAMM: model comparison (Table S11; inspect code for model specifications)

```{r dur-F0-gamm}

# Full model
matdurF0.gamm1 <- gamm4(mating_duration ~s(temp, k= 5, by= diet_fac, bs="cr") + diet_fac, random= ~ (1|replicate_vial_sep), data= mb.f0.analysis)

appraise(matdurF0.gamm1$gam)

# Model without diet-specific smooth terms
matdurF0.gamm2 <- gamm4(mating_duration ~s(temp, k= 5, bs="cr") + diet_fac, 
			    random= ~ (1|replicate_vial_sep), data= mb.f0.analysis)

# Model without any diet effects
matdurF0.gamm3 <- gamm4(mating_duration ~s(temp, k= 5, bs="cr"), 
			    random= ~ (1|replicate_vial_sep), data= mb.f0.analysis)

# Model with linear temperature term only
matdurF0.gamm3b <- gamm4(mating_duration ~ temp, 
			    random= ~ (1|replicate_vial_sep), data= mb.f0.analysis)

# Null model
matdurF0.gamm.null <- gamm4(mating_duration ~ 1, 
			    random= ~ (1|replicate_vial_sep), data= mb.f0.analysis )

# AIC model table (Table S11)
model.sel(matdurF0.gamm1$mer, matdurF0.gamm2$mer, matdurF0.gamm3$mer, matdurF0.gamm3b$mer, matdurF0.gamm.null$mer, rank = AIC)

```

### F0 Mating success, GAMM: model comparison (Table S12; inspect code for model specifications)

```{r success-F0-gamm}

# Full model
matsucF0.gamm1 <- gamm4(factor(mated) ~s(temp, k= 5, by= diet_fac, bs="cr") + diet_fac, family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= mb.f0a )

# Model without diet-specific smooth terms
matsucF0.gamm2 <- gamm4(factor(mated) ~s(temp, k= 5, bs="cr") + diet_fac, 
			    family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= mb.f0a )

matsucF0.gamm2b <- gamm4(factor(mated) ~ temp*diet_fac, 
			    family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= mb.f0a )

# Model without any diet effects
matsucF0.gamm3 <- gamm4(factor(mated) ~s(temp, k= 5, bs="cr"), 
			    family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= mb.f0a )

# Model with linear temperature term only
matsucF0.gamm3b <- gamm4(factor(mated) ~ temp, 
			    family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= mb.f0a )

# Null model
matsucF0.gamm.null <- gamm4(factor(mated) ~ 1, 
			    family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= mb.f0a )

# AIC model table (Table S12)
model.sel(matsucF0.gamm1$mer, matsucF0.gamm2$mer, matsucF0.gamm2b$mer, matsucF0.gamm3$mer, matsucF0.gamm3b$mer, matsucF0.gamm.null$mer, rank = AIC)

```

## F1 MALE MATING BEHAVIOUR

## F1 Latency to mate

### Maternal effects on F1 offspring: Latency to mate (Table S13)

```{r lat-F1-mat}

# Full model
latF1mat.gamm1 <- gamm4(latency_to_mate ~s(parental_temp, k= 5, by= parental_diet_fac, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))
appraise(latF1mat.gamm1$gam)

latF1mat.gamm1log <- gamm4(log(latency_to_mate) ~s(parental_temp, k= 5, by= parental_diet_fac, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))
appraise(latF1mat.gamm1log$gam)

# Model without diet-specific smooth terms
latF1mat.gamm2 <- gamm4(log(latency_to_mate) ~ s(parental_temp, k= 5, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))

# Model without any diet effects
latF1mat.gamm3 <- gamm4(log(latency_to_mate) ~ s(parental_temp, k= 5, bs="cr"), random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))

# Model with linear temperature term only
latF1mat.gamm3b <- gamm4(log(latency_to_mate) ~ parental_temp, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))

# Null model
latF1mat.gamm.null <- gamm4(log(latency_to_mate) ~ 1, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))

# AIC model table (Table S13)
model.sel(latF1mat.gamm1log$mer, latF1mat.gamm2$mer, latF1mat.gamm3$mer, latF1mat.gamm3b$mer, latF1mat.gamm.null$mer, rank = AIC)

```

### Paternal effects on F1 offspring: Latency to mate (Table S16)

```{r lat-F1-pat}

# Full model, untransformed response
latF1pat.gamm0 <- gamm4(latency_to_mate ~s(parental_temp, k= 5, by= parental_diet_fac, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

appraise(latF1pat.gamm0$gam) 

# Full model, log-transformed response
latF1pat.gamm1 <- gamm4(log(latency_to_mate) ~s(parental_temp, k= 5, by= parental_diet_fac, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

appraise(latF1pat.gamm1$gam)

# Model without diet-specific smooth terms
latF1pat.gamm2 <- gamm4(log(latency_to_mate) ~ s(parental_temp, k= 5, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

# Model without any diet effects
latF1pat.gamm3 <- gamm4(log(latency_to_mate) ~ s(parental_temp, k= 5, bs="cr"), random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

# Model with linear temperature term only
latF1pat.gamm3b <- gamm4(log(latency_to_mate) ~ parental_temp, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

# Null model
latF1pat.gamm.null <- gamm4(log(latency_to_mate) ~ 1, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

# AIC model table (Table S16)
model.sel(latF1pat.gamm1$mer, latF1pat.gamm2$mer, latF1pat.gamm3$mer, latF1pat.gamm3b$mer, latF1pat.gamm.null$mer, rank = AIC)

```

## F1 Mating duration

### F1 Mating duration, maternal effects (Table S14)

```{r dur-F1-mat}

# Full model
durF1mat.gamm1 <- gamm4(mating_duration ~s(parental_temp, k= 5, by= parental_diet_fac, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))

draw(durF1mat.gamm1$gam)
appraise(durF1mat.gamm1$gam)

# Model without diet-specific smooth terms
durF1mat.gamm2 <- gamm4(mating_duration ~ s(parental_temp, k= 5, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))

# Model without any diet effects
durF1mat.gamm3 <- gamm4(mating_duration ~ s(parental_temp, k= 5, bs="cr"), random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))

# Model with linear temperature term only
durF1mat.gamm3b <- gamm4(mating_duration ~ parental_temp, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))

# Null model 
durF1mat.gamm.null <- gamm4(mating_duration ~ 1, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "mat"))

# AIC model table (Table S14)
model.sel(durF1mat.gamm1$mer, durF1mat.gamm2$mer, durF1mat.gamm3$mer, durF1mat.gamm3b$mer, durF1mat.gamm.null$mer, rank = AIC)

```

### F1 Mating duration, paternal effects (Table S17)

```{r dur-F1-pat}

# Full model, untransformed response
durF1pat.gamm1 <- gamm4(mating_duration ~s(parental_temp, k= 5, by= parental_diet_fac, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

appraise(durF1pat.gamm1$gam) 

# Model without diet-specific smooth terms
durF1pat.gamm2 <- gamm4(mating_duration ~ s(parental_temp, k= 5, bs="cr") + parental_diet_fac, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

# Model without any diet effects
durF1pat.gamm3 <- gamm4(mating_duration ~ s(parental_temp, k= 5, bs="cr"), random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

# Model with linear temperature term only
durF1pat.gamm3b <- gamm4(mating_duration ~ parental_temp, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

# Null model
durF1pat.gamm.null <- gamm4(mating_duration ~ 1, random= ~ (1|replicate_vial_sep), data= subset(mb.f1.analysis, parental_effects== "pat"))

# AIC model table (Table S17)
model.sel(durF1pat.gamm1$mer, durF1pat.gamm2$mer, durF1pat.gamm3$mer, durF1pat.gamm3b$mer, durF1pat.gamm.null$mer, rank = AIC)

summary(durF1pat.gamm3$gam)
```

### Plot Figure 6: Male mating duration F1, paternal effects

```{r Figure-6}

# Create new data for predictions
x_durF1dat <- data.frame(parental_temp = seq(21, 29, length.out = 100))
pred_durF1patdat <- predict(durF1pat.gamm3$gam, x_durF1dat, se.fit = TRUE)

new_durF1patdat <- cbind(x_durF1dat, pred_durF1patdat)

# Calculate CIs 
new_durF1patdat <- data.frame(new_durF1patdat,
                      mu   = new_durF1patdat$fit,
                      lower  = new_durF1patdat$fit - 1.96 * new_durF1patdat$se.fit,
                      higher = new_durF1patdat$fit + 1.96 * new_durF1patdat$se.fit)

# Plot Figure 5

predDurF1pat_plot <- ggplot(new_durF1patdat) +
  geom_point(data= subset(mb.f1.analysis, parental_effects== "pat"), aes(x = parental_temp, y = mating_duration, color = factor(parental_diet), shape  = factor(parental_diet)), size = 3.2, alpha = 0.6, position= position_jitterdodge(dodge.width = 0.8, jitter.width = 0.1), show.legend = TRUE) +
  geom_smooth(aes(ymin = lower, ymax = higher, x = parental_temp, y = fit), stat = "identity", alpha = 0.1, color = "black", size = 1.1, show.legend = FALSE) +
  scale_x_continuous(breaks=c(21, 23, 25, 27, 29)) + 
  scale_color_manual(name  ="Paternal diet", values=c("#F8766D", "#619CFF", "#00BA38"), labels=c("restricted", "standard", "rich")) +
  scale_fill_manual(name  ="Paternal diet", values=c("#F8766D", "#619CFF", "#00BA38"), labels=c("restricted", "standard", "rich"))  +
  scale_shape_discrete(name  ="Paternal diet", labels=c("restricted", "standard", "rich")) + 
 # scale_color_manual(values=c("#F8766D", "#619CFF", "#00BA38")) +
 # scale_fill_manual(values=c("#F8766D", "#619CFF", "#00BA38")) +
  labs(title = "", color = "Paternal diet", y = "Mating duration [min]", x = "Temperature [degree Celsius]") + theme_classic(base_size = 23) 

predDurF1pat_plot

```

## F1 Male mating success

### F1 Mating success, maternal effects (Table S15)

```{r success-F1-mat}

matsucF1mat.gamm1 <- gamm4(factor(mated) ~s(parental_temp, k= 5, by= parental_diet_fac) + parental_diet_fac, family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "mat") )

appraise(latF1mat.gamm1$gam)

# Model without diet-specific smooth terms
matsucF1mat.gamm2 <- gamm4(factor(mated) ~s(parental_temp, k= 5) + parental_diet_fac, family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "mat") )

# Model without any diet effects
matsucF1mat.gamm3 <- gamm4(factor(mated) ~s(parental_temp, k= 5), family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "mat") )

# Model with linear temperature term only
matsucF1mat.gamm4 <- gamm4(factor(mated) ~ parental_temp, family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "mat") )

# Null model
matsucF1mat.gamm5 <- gamm4(factor(mated) ~ 1, family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "mat") )

# AIC model table (Table S15)
model.sel(matsucF1mat.gamm1$mer, matsucF1mat.gamm2$mer, matsucF1mat.gamm3$mer, matsucF1mat.gamm4$mer, matsucF1mat.gamm5$mer, rank = AIC)

```

### F1 Mating success, paternal effects (Table S18)

```{r success-F1-pat}

matsucF1pat.gamm1 <- gamm4(factor(mated) ~s(parental_temp, k= 5, by= parental_diet_fac) + parental_diet_fac, family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "pat") )

appraise(latF1pat.gamm1$gam)

# Model without diet-specific smooth terms
matsucF1pat.gamm2 <- gamm4(factor(mated) ~s(parental_temp, k= 5) + parental_diet_fac, family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "pat") )

# Model without any diet effects
matsucF1pat.gamm3 <- gamm4(factor(mated) ~s(parental_temp, k= 5), family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "pat") )

# Model with linear temperature term only
matsucF1pat.gamm4 <- gamm4(factor(mated) ~ parental_temp, family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "pat") )

# Null model
matsucF1pat.gamm5 <- gamm4(factor(mated) ~ 1, family= binomial, random= ~ (1|replicate_vial_sep) + (1|indobs), data= subset(mb.f1a, parental_effects== "pat") )

# AIC model table (Table S18)
model.sel(matsucF1pat.gamm1$mer, matsucF1pat.gamm2$mer, matsucF1pat.gamm3$mer, matsucF1pat.gamm4$mer, matsucF1pat.gamm5$mer, rank = AIC)

summary(matsucF1pat.gamm3$gam)

```

## F2 MALE MATING BEHAVIOUR

### F2 Latency to mate

```{r F2_latdat}

mb.f2.analysis <- mb.f2a %>% 
  filter(!is.na(latency_to_mate))
str(mb.f2.analysis)

```

### F2 Latency to mate, grandmaternal effects: LMER (Table S19)

```{r F2_lat_analysis_a}

latencyF2mat.lmer1 <- lmer(latency_to_mate ~ parental_temp*parental_diet_fac + I(parental_temp^2)*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4))) 
anova(latencyF2mat.lmer1)

latencyF2mat.lmer2 <- lmer(latency_to_mate ~ parental_temp + parental_diet_fac + I(parental_temp^2) + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4))) 
anova(latencyF2mat.lmer2)

latencyF2mat.lmer3 <- lmer(latency_to_mate ~ parental_temp  + I(parental_temp^2) + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4))) 

# Table S19
summary(latencyF2mat.lmer3, correlation = FALSE)

# Diagnostics
plot(simulateResiduals(latencyF2mat.lmer3))

```

### Plot preparation: F2 Latency to mate, grandmaternal effects

```{r F2_latmat_plot_a}

latF2.mat <- subset(mb.f2.analysis, parental_effects== "mat")

latencyF2mat.lmer3 <- lmer(latency_to_mate ~ parental_temp + I(parental_temp^2) + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4))) 
eF2mat <- allEffects(latencyF2mat.lmer3, xlevels= 300) 

```

### Plot Figure 7: F2 Latency to mate, grandmaternal effects

```{r Figure-7}

predMatLatF2_plot <- ggplot(as.data.frame(eF2mat)$parental_temp, aes(x= parental_temp, y = fit)) +
	geom_ribbon(aes(ymin = lower, ymax = upper), alpha= 0.1, linetype = 0, color = "black", size = 1.1) +
	geom_line(size= 1.2) +
	geom_point(data= mb.f2.analysis, aes(x= parental_temp, y= latency_to_mate, color = factor(parental_diet), shape = factor(parental_diet)), size = 3.2, alpha = 0.6, position= position_jitterdodge(dodge.width = 0.6, jitter.width = 0.1), show.legend = TRUE) + 
	xlab("Temperature [degree Celsius]")  +
	ylab("Latency to mate [min]")  + 
	scale_x_continuous(breaks= c(21, 23, 25, 27, 29)) +
	theme_bw(base_size = 23) +
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
		 panel.border = element_blank()) +
		 	scale_fill_manual(values= c("#F8766D", "#00BA38"), 
		 	                  name= "Grandmaternal\ndiet", labels= c("restricted", "rich")) +
	scale_colour_manual(values= c("#F8766D", "#00BA38"), 
                       name= "Grandmaternal\ndiet",
                       labels= c("restricted", "rich"))  +
	scale_shape_manual(values= c(16, 15), 
                       name= "Grandmaternal\ndiet",
                       labels= c("restricted", "rich")) 
	
predMatLatF2_plot

```

### F2 Latency to mate, grandpaternal effects (Table S21)

```{r F2_latency_pat}

latencyF2pat.lmer1 <- lmer(latency_to_mate ~ parental_temp*parental_diet_fac + I(parental_temp^2)*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "pat"), REML = T) #, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

# Table S21
summary(latencyF2pat.lmer1,  correlation = FALSE)

latencyF2pat.lmer2 <- lmer(latency_to_mate ~ parental_temp*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "pat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))
anova(latencyF2pat.lmer2)

anova(latencyF2pat.lmer1, latencyF2pat.lmer2)

latencyF2pat.lmer3 <- lmer(latency_to_mate ~ parental_temp + parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "pat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))
anova(latencyF2pat.lmer3)

latencyF2pat.lmer4 <- lmer(latency_to_mate ~ parental_temp + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "pat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))
anova(latencyF2pat.lmer4)

latencyF2pat.lmer5 <- lmer(latency_to_mate ~ parental_temp + I(parental_temp^2) + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "pat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))
anova(latencyF2pat.lmer5)

latencyF2pat.lmer.null <- lmer(latency_to_mate ~ 1 + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "pat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

anova(latencyF2pat.lmer4, latencyF2pat.lmer.null)

```

## F2 Mating duration

### F2 mating duration, grandmaternal effects (Table S22)

```{r F2_matdur_mat_}

matdurF2mat.lmer1 <- lmer(mating_duration ~ parental_temp*parental_diet_fac + I(parental_temp^2)*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

# Table S22
summary(matdurF2mat.lmer1, correlation = FALSE)

# Diagnostics
plot(simulateResiduals(matdurF2mat.lmer1))

matdurF2mat.lmer2 <- lmer(mating_duration ~ parental_temp*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))
anova(matdurF2mat.lmer2)

matdurF2mat.lmer3 <- lmer(mating_duration ~ parental_temp + parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))
anova(matdurF2mat.lmer3)

matdurF2mat.lmer4 <- lmer(mating_duration ~ parental_temp + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))
anova(matdurF2mat.lmer4)

```

## F2 mating duration, grandpaternal effects (Table S20)

```{r F2_matdur_pat_}

matdurF2pat.lmer1 <- lmer(mating_duration ~ parental_temp*parental_diet_fac + I(parental_temp^2)*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "pat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

# Table S20
summary(matdurF2pat.lmer1, correlation = FALSE) 

# Diagnostics
plot(simulateResiduals(matdurF2pat.lmer1))

matdurF2pat.lmer2 <- lmer(mating_duration ~ parental_temp + I(parental_temp^2) +
  parental_temp*parental_diet_fac +
  (1|replicate_vial_sep), data= subset(mb.f2.analysis, parental_effects== "pat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

anova(matdurF2pat.lmer1, matdurF2pat.lmer2)

```

### Plot Figure 8: F2 Mating duration, grandpaternal effects

```{r Figure-8)}

# Plot preparation: F2 Mating duration, paternal effects
matdurF2.pat <- subset(mb.f2.analysis, parental_effects== "pat")

matdurF2pat.lmer1 <- lmer(mating_duration ~ parental_temp*parental_diet_fac + I(parental_temp^2)*parental_diet_fac + (1|replicate_vial), data= matdurF2.pat, REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

pp <- 50
plot.df.matdurF2pat = data.frame( parental_temp = rep(seq(min(matdurF2.pat$parental_temp), max(matdurF2.pat$parental_temp),
       length.out= pp), 2),
       parental_diet_fac = rep(levels(matdurF2.pat$parental_diet_fac),times= 1, each= pp), mating_duration= 0 )

eF2pat <- as.data.frame(predict(matdurF2pat.lmer1, newdata= plot.df.matdurF2pat, re.form = NA))

plot.df.matdurF2pat$pred <- eF2pat[,1]
mm <- model.matrix(terms(matdurF2pat.lmer1), plot.df.matdurF2pat)
pvar1 <- diag(mm %*% tcrossprod(vcov(matdurF2pat.lmer1),mm))
cmult <- 1.96
plot.df.matdurF2pat <- data.frame(
    plot.df.matdurF2pat
    , lwr = plot.df.matdurF2pat$pred - cmult*sqrt(pvar1)
    , upr = plot.df.matdurF2pat$pred + cmult*sqrt(pvar1) )

#adding mean and SE of raw data
matdurF2.pat.means <- mb.f2.analysis %>%
  filter(parental_effects == "pat") %>%
  group_by(parental_temp, parental_diet_fac) %>%
  summarise(mean_latency_to_mate = mean(latency_to_mate, na.rm = TRUE),
            sd_latency_to_mate = sd(latency_to_mate),
            n_latency_to_mate = n(),
            SE_latency_to_mate = sd(latency_to_mate)/sqrt(n()),
		mean_mating_duration = mean(mating_duration, na.rm = TRUE),
            sd_mating_duration = sd(mating_duration),
            n_mating_duration = n(),
            SE_mating_duration = sd(mating_duration)/sqrt(n()))

# Figure 8

matdur.F2.pat.plot <- ggplot(plot.df.matdurF2pat, aes(x= parental_temp, y = pred, col= parental_diet_fac, fill = parental_diet_fac)) +
	geom_ribbon(aes(ymin = lwr, ymax = upr, col= parental_diet_fac), alpha= 0.1, linetype = 0) +
	geom_smooth(method = "lm", formula = y ~ x + I(x^2), aes(col= parental_diet_fac, linetype= parental_diet_fac), size= 1.5, se = F) +
	geom_point(data= subset(mb.f2.analysis, parental_effects== "pat"),
	  aes(x= parental_temp, y= mating_duration, col= parental_diet_fac, shape = parental_diet_fac), size = 3.2, alpha = 0.6, position= position_jitterdodge(dodge.width = 0.6, jitter.width = 0.1), show.legend = TRUE) +
	xlab("Temperature [degree Celsius]")  +
	ylab("Mating duration [min]")  +
	#ggtitle("(b) Grand-paternal effects in F2 males:\nPredicted mating duration") +
	scale_x_continuous(breaks= c(21, 23, 25, 27, 29)) +
	theme_bw(base_size = 23) +
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
		 panel.border = element_blank()) +
	scale_fill_manual(name= "Grandpaternal\ndiet", labels= c("restricted", "rich"),
			     values= c("#F8766D", "#00BA38")) +
	scale_colour_manual(values= c("#F8766D", "#00BA38"),
                       name= "Grandpaternal\ndiet",
                       labels= c("restricted", "rich")) +
	scale_shape_manual(values= c(16, 15),
                       name= "Grandpaternal\ndiet",
                       labels= c("restricted", "rich")) +
  scale_linetype_manual(name  ="Grandpaternal\ndiet", values = c("solid", "dotted"), labels=c("restricted", "rich"))

matdur.F2.pat.plot

```

### F2 Mating success, grandmaternal effects (Table S23)

```{r F2 matsuc_mat}

# GLMER not converging
matsucF2mat.lmer1.04 <- glmer(factor(mated) ~ parental_temp + I(parental_temp^2) + (1|indobs), family= binomial, data= subset(mb.f2a, parental_effects== "mat"))

# GLM
matsucF2mat.glm1 <- glm(factor(mated) ~ parental_diet_fac*parental_temp + parental_diet_fac*I(parental_temp^2), family= quasibinomial, data= subset(mb.f2a, parental_effects== "mat"))

# Table S23
summary(matsucF2mat.glm1)

matsucF2mat.glm2 <- glm(factor(mated) ~ parental_diet_fac*parental_temp, family= quasibinomial, data= subset(mb.f2a, parental_effects== "mat"))
summary(matsucF2mat.glm2)

matsucF2mat.glm3 <- glm(factor(mated) ~ parental_temp + I(parental_temp^2), family= quasibinomial, data= subset(mb.f2a, parental_effects== "mat"))

anova(matsucF2mat.glm1, matsucF2mat.glm2, test= "Chisq")

anova(matsucF2mat.glm1, matsucF2mat.glm3, test= "Chisq")

anova(matsucF2mat.glm2, matsucF2mat.glm3, test= "Chisq")

matsucF2mat.glm4 <- glm(factor(mated) ~ parental_diet_fac+parental_temp, family= quasibinomial, data= subset(mb.f2a, parental_effects== "mat"))
anova(matsucF2mat.glm2, matsucF2mat.glm4, test= "Chisq")

matsucF2mat.glm5 <- glm(factor(mated) ~ parental_diet_fac, family= quasibinomial, data= subset(mb.f2a, parental_effects== "mat"))
anova(matsucF2mat.glm4, matsucF2mat.glm5, test= "Chisq")

matsucF2mat.glm6 <- glm(factor(mated) ~ parental_temp, family= quasibinomial, data= subset(mb.f2a, parental_effects== "mat"))
anova(matsucF2mat.glm4, matsucF2mat.glm6, test= "Chisq")

matsucF2mat.glm.null <- glm(factor(mated) ~ 1, family= quasibinomial, data= subset(mb.f2a, parental_effects== "mat"))

anova(matsucF2mat.glm5, matsucF2mat.glm.null, test= "Chisq")

anova(matsucF2mat.glm6, matsucF2mat.glm.null, test= "Chisq")

```

### F2 Mating success, grandpaternal effects (Table S24)

```{r F2 matsuc_pat}

# GLM
matsucF2pat.glm1 <- glm(factor(mated) ~ parental_diet_fac*parental_temp + parental_diet_fac*I(parental_temp^2), family= quasibinomial, data= subset(mb.f2a, parental_effects== "pat"))

# Table S24
summary(matsucF2pat.glm1)

matsucF2pat.glm2 <- glm(factor(mated) ~ parental_diet_fac*parental_temp, family= quasibinomial, data= subset(mb.f2a, parental_effects== "pat"))
summary(matsucF2pat.glm2)

matsucF2pat.glm3 <- glm(factor(mated) ~ parental_temp + I(parental_temp^2), family= quasibinomial, data= subset(mb.f2a, parental_effects== "pat"))

anova(matsucF2pat.glm1, matsucF2pat.glm2, test= "Chisq")

anova(matsucF2pat.glm1, matsucF2pat.glm3, test= "Chisq")

anova(matsucF2pat.glm2, matsucF2pat.glm3, test= "Chisq")

matsucF2pat.glm4 <- glm(factor(mated) ~ parental_diet_fac+parental_temp, family= quasibinomial, data= subset(mb.f2a, parental_effects== "pat"))

anova(matsucF2pat.glm2, matsucF2pat.glm4, test= "Chisq")

matsucF2pat.glm5 <- glm(factor(mated) ~ parental_temp, family= quasibinomial, data= subset(mb.f2a, parental_effects== "pat"))

anova(matsucF2pat.glm4, matsucF2pat.glm5, test= "Chisq")

matsucF2pat.glm6 <- glm(factor(mated) ~ parental_diet_fac, family= quasibinomial, data= subset(mb.f2a, parental_effects== "pat"))

anova(matsucF2pat.glm4, matsucF2pat.glm6, test= "Chisq")

matsucF2pat.glm.null <- glm(factor(mated) ~ 1, family= quasibinomial, data= subset(mb.f2a, parental_effects== "pat"))
anova(matsucF2pat.glm5, matsucF2pat.glm.null, test= "Chisq")

anova(matsucF2pat.glm6, matsucF2pat.glm.null, test= "Chisq")

```

## F3 MALE MATING BEHAVIOUR

### F3 Latency to mate

```{r F3_a}

mb.f3.analysis <- mb.f3a %>%
  filter(!is.na(latency_to_mate))
str(mb.f3.analysis)
```

### F3 Latency to mate, great-grandmaternal and great-grandpaternal effects (Tables S25, S26)

```{r F3_latency_maternal_paternal}

# F3 Latency to mate, maternal effects

latencyF3mat.lmer1 <- lmer(latency_to_mate ~ parental_temp*parental_diet_fac + I(parental_temp^2)*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f3.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

# Table S25
summary(latencyF3mat.lmer1, correlation = FALSE)

# F3 Latency to mate, paternal effects

latencyF3pat.lmer1 <- lmer(latency_to_mate ~ parental_temp*parental_diet_fac + I(parental_temp^2)*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f3.analysis, parental_effects== "pat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

# Table S26
summary(latencyF3pat.lmer1, correlation = FALSE)

```

### F3 Mating duration, great-grandmaternal effects (Tables S27, S28)

```{r F3_matingduration_maternal_paternal}

matdurF3mat.lmer1 <- lmer(mating_duration ~ parental_temp*parental_diet_fac + I(parental_temp^2)*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f3.analysis, parental_effects== "mat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

# Table S27
summary(matdurF3mat.lmer1, correlation = FALSE)

# F3 Mating duration, paternal effects

matdurF3pat.lmer1 <- lmer(mating_duration ~ parental_temp*parental_diet_fac + I(parental_temp^2)*parental_diet_fac + (1|replicate_vial_sep), data= subset(mb.f3.analysis, parental_effects== "pat"), REML = T, control= lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4)))

# Table S28
summary(matdurF3pat.lmer1, correlation = FALSE)

```

## F3 Mating success

```{r F3_matingsuccess_a}
mb.f3.analysis2 <- mb.f3a
str(mb.f3.analysis2)
```

### F3 Mating success, great-grandmaternal and great-grandpaternal effects (Tables S29, S30)

```{r F3_matingsuc_mat_pat}

# GLMER not converging
matsucF3mat.lmer1 <- glmer(factor(mated) ~ parental_temp + I(parental_temp^2) + (1|indobs), family= binomial, data= subset(mb.f3.analysis2, parental_effects== "mat" & parental_diet_fac== "1.5" ))

# GLM
matsucF3mat.glm1 <- glm(factor(mated) ~ parental_diet_fac*parental_temp + parental_diet_fac*I(parental_temp^2), family= quasibinomial, data= subset(mb.f3.analysis2, parental_effects== "mat"))

# Table S29
summary(matsucF3mat.glm1) 

## F3 Mating success, paternal effects

# GLM
matsucF3pat.glm1 <- glm(factor(mated) ~ parental_diet_fac*parental_temp + parental_diet_fac*I(parental_temp^2), family= quasibinomial, data= subset(mb.f3.analysis2, parental_effects== "pat"))

# Table S30
summary(matsucF3pat.glm1)

```
