---
title: "Survival analyses"
author: "Zajitschek et al."
date: "2023-06-01"
output: 
  workflowr::wflow_html:
   code_folding: hide
editor_options: 
  chunk_output_type: console
---

## Load libraries and helper function
```{r load-libraries, message=FALSE, warning=FALSE, results="hide"}
library(readr)      # read data
library(dplyr)      # data structure
library(reshape2)   # data structure
library(tidyverse)  # data structure
library(rcompanion) # mean and median (+- bootstrapped CI)
library(survival)   # survival analysis
library(coxme)      # mixed-effects Cox proportional hazard model
library(lme4)       # generalized linear mixed-effects model
library(mgcv)       # generalized additive models for Cox regression
library(MuMIn)      # AIC results table
library(ggplot2)    # plot
library(survminer)  # plot
library(sjPlot)     # plot
library(kableExtra) # table formatting
library(splines)    # spline functions

kable.table <- function(df) {
  kable(df, "html") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px", box_css = "border: 0px;")
}
```

## Load survival data (F0 and F1)
```{r import-survival, message=FALSE}

#F0 survival data
survival <- read_csv("./data/survival.csv")
survival$diet <- as.factor(survival$diet)
survival$vial_id <- as.factor(survival$vial_id)
survival$rep <- as.factor(survival$rep)

#F1 survival data
survivalF1 <- read_csv("./data/survivalF1.csv")
survivalF1$par_diet <- as.factor(survivalF1$par_diet)
survivalF1$vial_id <- as.factor(survivalF1$vial_id)
survivalF1$rep <- as.factor(survivalF1$rep)

```

## F0 SURVIVAL

### Calculate F0 mean lifespan with bootstrapped CI
```{r F0-mean-lifespan-summary, message=FALSE}

# plot(survival$lifespan) # ok: no off-values/outliers
# max(survival$lifespan)  # max. lifespan = 152 days

# Bootstrap replicates for CI
boot.ls <- groupwiseMean(lifespan ~ sex*temperature*diet, data = survival, boot = T, bca = T) 
medians.F1 <- survival %>% group_by(temperature, diet, sex) %>%
  summarise(., mean = mean(lifespan), se = sd(lifespan)/sqrt(length(lifespan)), sd = sd(lifespan), 
   median = median(lifespan), n_vials = n_distinct(vial_id), n_inds = n()) %>%
  dplyr::select(sex, median, n_vials, n_inds)

boot.ls2 <- left_join(boot.ls, medians.F1) %>%
  dplyr::select(- c(n, Mean, Conf.level:Trad.upper))

kable.table(boot.ls2)

```

### Plot F0 mean lifespan for males and females (mean +- 95% CI) with diet (a, b) or temperature (c, d) on x-axis (not in manuscript)

```{r plot-F0-mean-lifespan}

mycols <- colors()[c(30, 62, 119, 504, 554)]

# 1) Diet on x-, lifespan on y-axis

plot.meanlifespan.m <- ggplot(data= subset(boot.ls, sex== "male"), aes(x= as.factor(diet), y= Boot.mean, fill= as.factor(temperature))) +  
	scale_fill_manual(values= mycols, name = "Temperature") +
     	geom_bar(position=position_dodge(), stat = "identity", colour="black") + 
	geom_errorbar(aes(ymin= Bca.lower, ymax= Bca.upper), width=.4, colour="black",
	position=position_dodge(0.9)) +
  expand_limits(y = 120) +
	ylab("Lifespan (days)")  + 
	xlab("Diet")  + 
	ggtitle("Male lifespan") + 
	theme_bw(base_size = 18)+
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
	panel.border = element_blank(),  legend.position = "none",
	axis.line.x = element_line(), axis.line.y = element_line()) 

plot.meanlifespan.f <- ggplot(data= subset(boot.ls, sex== "female"), aes(x= as.factor(diet), y= Boot.mean, fill= as.factor(temperature))) +  
	scale_fill_manual(values= mycols, name = "Temp") +
     	geom_bar(position=position_dodge(), stat = "identity", colour="black") + 
	#geom_point(size = 3.5, position=position_dodge(0.1)) + 	
	geom_errorbar(aes(ymin= Bca.lower, ymax= Bca.upper), width=.4, colour="black",
	position=position_dodge(0.9)) +
  expand_limits(y = 120) +
	ylab("Lifespan (days)")  + 
	xlab("Diet")  + 
	ggtitle("Female lifespan") + 
	theme_bw(base_size = 18)+
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
	panel.border = element_blank(), legend.position = c(.89, .9), legend.text = element_text(size = 10), legend.title = element_text(size = 10), legend.key.size = unit(0.6, "lines"), 
	axis.line.x = element_line(), axis.line.y = element_line()) 

ggarrange(plot.meanlifespan.m, plot.meanlifespan.f,
  labels = c("a", "b"),
  ncol = 2, nrow = 1)

# 2) Temperature on x-, lifespan on y-axis

plot.meanlifespan2.m <- ggplot(data= subset(boot.ls, sex== "male"), aes(x= as.factor(temperature), y= Boot.mean, fill= as.factor(diet))) +  
	scale_fill_manual(values= mycols, name = "Diet") +
     	geom_bar(position=position_dodge(), stat = "identity", colour="black") + 
	geom_errorbar(aes(ymin= Bca.lower, ymax= Bca.upper), width=.4, colour="black",
	position=position_dodge(0.9)) +
  expand_limits(y = 120) +
	ylab("Lifespan (days)")  + 
	xlab("Temperature")  + 
	ggtitle("Male lifespan") + 
	theme_bw(base_size = 18)+
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
	panel.border = element_blank(),  legend.position = "none",
	axis.line.x = element_line(), axis.line.y = element_line()) 

plot.meanlifespan2.f <- ggplot(data= subset(boot.ls, sex== "female"), aes(x= as.factor(temperature), y= Boot.mean, fill= as.factor(diet))) +  
	scale_fill_manual(values= mycols, name = "Diet") +
     	geom_bar(position=position_dodge(), stat = "identity", colour="black") + 
	geom_errorbar(aes(ymin= Bca.lower, ymax= Bca.upper), width=.4, colour="black",
	position=position_dodge(0.9)) +
  expand_limits(y = 120) +
	ylab("Lifespan (days)")  + 
	xlab("Temperature")  + 
	ggtitle("Female lifespan") + 
	theme_bw(base_size = 18)+
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
	panel.border = element_blank(), legend.position = c(.8, .8), legend.text = element_text(size = 10), legend.title = element_text(size = 10), legend.key.size = unit(0.6, "lines"),
	axis.line.x = element_line(), axis.line.y = element_line()) 

ggarrange(plot.meanlifespan2.m, plot.meanlifespan2.f,
  labels = c("c", "d"),
  ncol = 2, nrow = 1)
```

### Plot Figure 1: Survival curves (separate plots per temperature and sex)

```{r F0-survival-curves-fig, fig.show = "hold", out.width = "70%"}

survival$groups <- interaction(survival$sex, survival$temperature, survival$diet)

surv.m21 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "21"))
surv.m23 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "23"))
surv.m25 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "25"))
surv.m27 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "27"))
surv.m29 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "29"))
surv.f21 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "21"))
surv.f23 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "23"))
surv.f25 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "25"))
surv.f27 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "27"))
surv.f29 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "29"))

# Set upper lifespan limit for all survival plots
xmax <- 154 # max(survival$lifespan) + 2  

ggsurvplot(surv.m21, title = "(a) Males at 21 degree Celsius", 
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", data = subset(survival, sex=="male" & temperature == "21"),
  legend.labs = c("restricted", "standard", "rich"))

ggsurvplot(surv.m23, title = "(c) males at 23 degree Celsius", 
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", data = subset(survival, sex=="male" & temperature == "23"),
  legend.labs = c("restricted", "standard", "rich"))

ggsurvplot(surv.m25, title = "(e) males at 25 degree Celsius", 
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", data = subset(survival, sex=="male" & temperature == "25"),
  legend.labs = c("restricted", "standard", "rich"))

ggsurvplot(surv.m27, title = "(g) males at 27 degree Celsius",
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", data = subset(survival, sex=="male" & temperature == "27"),
  legend.labs = c("restricted", "standard", "rich"))

ggsurvplot(surv.m29, title = "(i) males at 29 degree Celsius",
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", data = subset(survival, sex=="male" & temperature == "29"),
  legend.labs = c("restricted", "standard", "rich"))

ggsurvplot(surv.f21, title = "(b) females at 21 degree Celsius",
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", legend.labs = c("restricted", "standard", "rich"),
  data = subset(survival, sex=="female" & temperature == "21"))

ggsurvplot(surv.f23, title = "(d) females at 23 degree Celsius",
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", legend.labs = c("restricted", "standard", "rich"),
  data = subset(survival, sex=="female" & temperature == "23"))

ggsurvplot(surv.f25, title = "(f) females at 25 degree Celsius", 
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", legend.labs = c("restricted", "standard", "rich"),
  data = subset(survival, sex=="female" & temperature == "25"))

ggsurvplot(surv.f27, title = "(h) females at 27 degree Celsius", 
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", legend.labs = c("restricted", "standard", "rich"),
  data = subset(survival, sex=="female" & temperature == "27"))

ggsurvplot(surv.f29, title = "(j) females at 29 degree Celsius", 
  xlim = c(0, xmax), xlab = "Age (days)", break.time.by = 20, 
  risk.table = F, conf.int = T, pval = F, legend = "right", 
  legend.title = "", legend.labs = c("restricted", "standard", "rich"),
  data = subset(survival, sex=="female" & temperature == "29"))

```

```{r survival-curves-fig1, eval=FALSE, include=FALSE}

survival$groups <- interaction(survival$sex, survival$temperature, survival$diet)

surv.m21 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "21"))
surv.m23 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "23"))
surv.m25 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "25"))
surv.m27 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "27"))
surv.m29 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "male" & temperature == "29"))
surv.f21 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "21"))
surv.f23 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "23"))
surv.f25 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "25"))
surv.f27 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "27"))
surv.f29 <- survfit(Surv(lifespan, status) ~ groups, conf.type = "plain", data = subset(survival, sex == "female" & temperature == "29"))

#set colours
survcol <- c("#F8766D", "#619CFF", "#00BA38")
survline <- c( "dotted", "dashed", "solid")

xmax <- 154 # max(survival$lifespan) + 2  # set upper lifespan limit for all survival plots
splots <- list()
splots[[1]] <- ggsurvplot(surv.m21, title = "(a) males at 21 deg C", xlim = c(0, xmax), xlab = "", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,  palette = survcol, linetype = survline,
		pval = F, font.title=24, size=1.8, font.x=22, font.legend=22, font.y=22, legend = 0, font.tickslab=22, data = subset(survival, sex=="male" & temperature == "21"))
splots[[2]] <- ggsurvplot(surv.m23, title = "(c) males at 23 deg C", xlim = c(0, xmax), ylab = "", xlab = "", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,  palette = survcol,linetype = survline,
		pval = F, font.title=24,  size=1.8, font.x=22, font.legend=22, font.y=22, legend = 0, font.tickslab=22, data = subset(survival, sex=="male" & temperature == "23"))
splots[[3]] <- ggsurvplot(surv.m25, title = "(e) males at 25 deg C", xlim = c(0, xmax), ylab = "", xlab = "", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,  palette = survcol, linetype = survline,
		pval = F, font.title=24,  size=1.8, font.x=22, font.legend=22, font.y=22, legend = 0, font.tickslab=22, data = subset(survival, sex=="male" & temperature == "25"))
splots[[4]] <- ggsurvplot(surv.m27, title = "(g) males at 27 deg C", xlim = c(0, xmax), ylab = "", xlab = "", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,  palette = survcol, linetype = survline,
		pval = F, font.title=24,  size=1.8, font.x=22, font.legend=22, font.y=22, legend = 0, font.tickslab=22, data = subset(survival, sex=="male" & temperature == "27"))
splots[[5]] <- ggsurvplot(surv.m29, title = "(i) males at 29 deg C", xlim = c(0, xmax), ylab = "", xlab = "Age (days)", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,  palette = survcol, linetype = survline,
		pval = F, font.title=24,  size=1.8, font.x=22, font.legend=22, font.y=22, legend = 0, font.tickslab=22, data = subset(survival, sex=="male" & temperature == "29"))
splots[[6]] <- ggsurvplot(surv.f21, title = "(b) females at 21 deg C", xlim = c(0, xmax), xlab = "", ylab = "", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,  palette = survcol, linetype = survline,
		pval = F, font.title=24, size=1.8, font.x=22, font.legend=22, font.y=22, legend = 0, font.tickslab=22, data = subset(survival, sex=="female" & temperature == "21"))
splots[[7]] <- ggsurvplot(surv.f23, title = "(d) females at 23 deg C", xlim = c(0, xmax), ylab = "", xlab = "", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,  palette = survcol, linetype = survline,
		pval = F, font.title=24,  size=1.8, font.x=22, font.legend=22, font.y=22, legend = 0, font.tickslab=22, data = subset(survival, sex=="female" & temperature == "23"))
splots[[8]] <- ggsurvplot(surv.f25, title = "(f) females at 25 deg C", xlim = c(0, xmax), ylab = "", xlab = "", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,  palette = survcol, linetype = survline,
		pval = F, font.title=24,  size=1.8, font.x=22, font.legend=22, font.y=22, legend = 0, font.tickslab=22, data = subset(survival, sex=="female" & temperature == "25"))
splots[[9]] <- ggsurvplot(surv.f27, title = "(h) females at 27 deg C", xlim = c(0, xmax), ylab = "", xlab = "", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,  palette = survcol, linetype = survline,
		pval = F, font.title=24,  size=1.8, font.x=22, font.legend=22, font.y=22, legend = 0, font.tickslab=22, data = subset(survival, sex=="female" & temperature == "27"))
splots[[10]] <- ggsurvplot(surv.f29, title = "(j) females at 29 deg C", xlim = c(0, xmax), ylab = "", xlab = "Age (days)", break.time.by = 20, legend.title = "", risk.table = F, conf.int = T,
		pval = F, font.title=24,  size=1.8, font.x=22, font.legend=22, font.y=22, legend = c(0.8, 0.55), font.tickslab=22, data = subset(survival, sex=="female" & temperature == "29"),  palette = survcol, linetype = survline,
		legend.labs = c("restricted", "standard", "rich"))

# Arrange multiple ggsurvplots and print the output (Figure 1 in the manuscript)
plot.survivalcurves <- arrange_ggsurvplots(splots, print = T, ncol = 2, nrow = 5)
ggsave("survivalcurves_rev20-3-2023.pdf", plot.survivalcurves, width= 12, height= 15)

```

## F0 Survival analyses

### Data preparation F0 survival

```{r survival-model-data-prep-F0}

# Create grouping variables
survival$combined <- interaction(survival$sex, survival$diet, survival$temperature)
survival$specificvial <- interaction(survival$combined, survival$rep)

# Create subsets for each temperature
survival.21 <- subset(survival, temperature == "21")
survival.23 <- subset(survival, temperature == "23")
survival.25 <- subset(survival, temperature == "25")
survival.27 <- subset(survival, temperature == "27")
survival.29 <- subset(survival, temperature == "29")
```

## Sex-specific mixed Cox Proportional Hazards model F0

### 1) F0 females, Cox PH: proportionality assumption test
```{r coxme-female-F0}

# Model specification
cox.female <- coxme(Surv(lifespan, status) ~ diet*temperature + (1|specificvial), 
  data= subset(survival, sex == "female"))

#Cox proportionality assumption test: all effects p < 0.001 => violated
cox.zph(coxph(Surv(lifespan, status) ~ diet*temperature, data= subset(survival, sex == "female")))
```

## Discrete time survival models, since proportionality assumption is violated

### Data preparation and model specifications for males and females (inspect code for model specifications used in model selection tables below)

```{r survival-glmer, message=FALSE, warning=FALSE}

# Convert into long data format
survival$indspec <- with(survival, interaction(vial_id, ind))

survival_long <- survival %>% group_by(indspec) %>% 
  complete(lifespan = seq(1, max(lifespan)-1), nesting(vial_id, temperature, diet, sex, rep, ind, combined, specificvial), fill = list(status = 0)) 

survival_long$temp_fac <- as.factor(survival_long$temperature)

# Females
glmer_surv_fem_F0_1 <- glmer(status ~ diet * ns(temperature)  + (1|specificvial),family = binomial, data = subset(survival_long, sex == "female"), nAGQ = 0)

glmer_surv_fem_F0_2 <- glmer(status ~ diet + ns(temperature)  + (1|specificvial),family = "binomial", data = subset(survival_long, sex == "female"), nAGQ = 0)

glmer_surv_fem_F0_3 <- glmer(status ~ ns(temperature)  + (1|specificvial),family = "binomial", data = subset(survival_long, sex == "female"), nAGQ = 0)

glmer_surv_fem_F0_4 <- glmer(status ~ diet  + (1|specificvial),family = "binomial", data = subset(survival_long, sex == "female"), nAGQ = 0)

glmer_surv_fem_F0_5 <- glmer(status ~ 1  + (1|specificvial),family = "binomial", data = subset(survival_long, sex == "female"), nAGQ = 0)


# Males 
glmer_surv_mal_F0_1 <- glmer(status ~ diet * ns(temperature)  + (1|specificvial),family = "binomial", data = subset(survival_long, sex == "male"), nAGQ = 0)

glmer_surv_mal_F0_2 <- glmer(status ~ diet + ns(temperature)  + (1|specificvial),family = "binomial", data = subset(survival_long, sex == "male"), nAGQ = 0)

glmer_surv_mal_F0_3 <- glmer(status ~ ns(temperature)  + (1|specificvial),family = "binomial", data = subset(survival_long, sex == "male"), nAGQ = 0)

glmer_surv_mal_F0_4 <- glmer(status ~ diet  + (1|specificvial),family = "binomial", data = subset(survival_long, sex == "male"), nAGQ = 0)

glmer_surv_mal_F0_5 <- glmer(status ~ 1 + (1|specificvial), family = "binomial", data = subset(survival_long, sex == "male"), nAGQ = 0)

```

### Compare models for females, using AIC (Table S4)
```{r survival-F0-female_models}
model.sel(glmer_surv_fem_F0_1, glmer_surv_fem_F0_2 ,glmer_surv_fem_F0_3, glmer_surv_fem_F0_4, glmer_surv_fem_F0_5, rank = AIC)
```

### Compare models for males, using AIC (Table S4)
```{r survival-F0-male_models}
model.sel(glmer_surv_mal_F0_1, glmer_surv_mal_F0_2 ,glmer_surv_mal_F0_3, glmer_surv_mal_F0_4, glmer_surv_mal_F0_5, rank = AIC)
```

## Logrank tests F0 survival: Females (Table S5)

### 1) F0 females: Diet effect at 21 degree Celsius

```{r logrank-females-21degC}
f21 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.21, sex == "female"))
print(f21, digits = 3)
```

### 2) F0 females: Diet effect at 23 degree Celsius
```{r logrank-females-23degC}
f23 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.23, sex == "female"))
print(f23, digits = 3)
```

### 3) F0 females: Diet effect at 25 degree Celsius
```{r logrank-females-25degC}
f25 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.25, sex == "female"))
print(f25, digits = 3)
```

### 4) F0 females: Diet effect at 27 degree Celsius
```{r logrank-females-27degC}
f27 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.27, sex == "female"))
print(f27, digits = 3)
```

### 5) F0 females: Diet effect at 29 degree Celsius
```{r logrank-females-29degC}
f29 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.29, sex == "female"))
print(f29, digits = 3)
```

## Logrank tests F0 survival: Males (Table S5)

### 1) F0 males: Diet effect at 21 degree Celsius
```{r logrank-males-21degC}
m21 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.21, sex == "male"))
print(m21, digits = 3)
```

### 2) F0 males: Diet effect at 23 degree Celsius
```{r logrank-males-23degC}
m23 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.23, sex == "male"))
print(m23, digits = 3)
```

### 3) F0 males: Diet effect at 25 degree Celsius
```{r logrank-males-25degC}
m25 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.25, sex == "male"))
print(m25, digits = 3)
```

### 4) F0 males: Diet effect at 27 degree Celsius
```{r logrank-males-27degC}
m27 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.27, sex == "male"))
print(m27, digits = 3)
```

### 5) F0 males: Diet effect at 29 degree Celsius
```{r logrank-males-29degC}
m29 <- pairwise_survdiff(Surv(lifespan, status) ~ diet, data = subset(survival.29, sex == "male"))
print(m29, digits = 3)
```

## F1 SURVIVAL

### Explore F1 survival data

```{r explore-F1-survival, message=FALSE}

# plot(survivalF1$lifespan) # ok: no off-values/outliers
# max(survivalF1$lifespan)  # max. lifespan = 100 days

survivalF1 %>% group_by(par_temp, par_diet, sex) %>%
  summarise(., mean = mean(lifespan), se = sd(lifespan)/sqrt(length(lifespan)), sd = sd(lifespan), 
   median = median(lifespan), n_vials = n_distinct(vial_id), n_inds = n())

```

### Calculate F1 mean lifespan with bootstrapped CI

```{r mean-F1-lifespan-CI}

bootF1.ls <- groupwiseMean(lifespan ~ par_temp*par_diet*sex, data = survivalF1, boot = T, bca = T)  #5000 bootstrap replicates for CI
bootF1.ls

```

### Plot F1 mean lifespan for males and females (not in manuscript)

```{r plot-F1-mean-lifespan}

mycols <- colors()[c(30, 62, 119, 504, 554)]

plotF1.meanlifespan.m <- ggplot(data= subset(bootF1.ls, sex== "m"), aes(x= as.factor(par_diet), y= Boot.mean, fill= as.factor(par_temp))) +  
	#scale_fill_grey(start=0.8, end=0.3, name = "Temperature") +
	scale_fill_manual(values= mycols, name = "Parental\ntemperature") +
     	geom_bar(position=position_dodge(), stat = "identity", colour="black") + 
	#geom_point(size = 3.5, position=position_dodge(0.1)) + 	
	geom_errorbar(aes(ymin= Bca.lower, ymax= Bca.upper), width=.4, colour="black",
	position=position_dodge(0.9)) +
  expand_limits(y = 82) +
	ylab("Lifespan (days)")  + 
	xlab("Parental diet")  + 
	ggtitle("Male lifespan (mean \U00B1 95% CI)") + 
	theme_bw(base_size = 18)+
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
	panel.border = element_blank(), 
	axis.line.x = element_line(), axis.line.y = element_line()) 

plotF1.meanlifespan.f <- ggplot(data= subset(bootF1.ls, sex== "f"), aes(x= as.factor(par_diet), y= Boot.mean, fill= as.factor(par_temp))) +  
	#scale_fill_grey(start=0.8, end=0.3, name = "Temperature") +
	scale_fill_manual(values= mycols, name = "Parental\ntemperature") +
     	geom_bar(position=position_dodge(), stat = "identity", colour="black") + 
	#geom_point(size = 3.5, position=position_dodge(0.1)) + 	
	geom_errorbar(aes(ymin= Bca.lower, ymax= Bca.upper), width=.4, colour="black",
	position=position_dodge(0.9)) +
  expand_limits(y = 82) +
	ylab("Lifespan (days)")  + 
	xlab("Parental diet")  + 
	ggtitle("Female lifespan (mean \U00B1 95% CI)") + 
	theme_bw(base_size = 18)+
	theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
	panel.border = element_blank(), 
	axis.line.x = element_line(), axis.line.y = element_line()) 

plotF1.meanlifespan.m
plotF1.meanlifespan.f

```

## F1 Survival analyses

### Data preparation F1 survival

```{r F1-survival-model-data-prep}

# Create grouping variables
survivalF1$combined <- interaction(survivalF1$sex, survivalF1$par_diet, survivalF1$par_temp)
survivalF1$specificvial <- interaction(survivalF1$combined, survivalF1$rep)

```

## Sex-specific mixed Cox Proportional Hazards models F1

### 1) F1 females, Cox PH: Model summary (Table S6) and proportionality assumption test

```{r coxme-F1-females}

# Model specification: par_eff represents either maternal or paternal effects
coxF1.f1 <- coxme(Surv(lifespan, status) ~ par_eff*par_temp*par_diet + (1|specificvial), data = subset(survivalF1, sex == "f"))

# Model summary (custom function extracts 3 digit z- and p-value)
extract_coxme_table <- function (mod){
        beta <- mod$coefficients #$fixed is not needed
        nvar <- length(beta)
        nfrail <- nrow(mod$var) - nvar
        se <- sqrt(diag(mod$var)[nfrail + 1:nvar])
        z<- round(beta/se, 3)
        p<- signif(1 - pchisq((beta/se)^2, 1), 3)
        table=data.frame(cbind(z,p))
        return(table)
    }
#extract_coxme_table(coxF1.f1)
summary(coxF1.f1, digits = 3)

f1.zph <- coxph(Surv(lifespan, status) ~ par_eff*par_temp*par_diet, data = survivalF1)
cox.zph(coxph(Surv(lifespan, status) ~ par_eff*par_temp*par_diet, data = survivalF1))
```

### Female F1 survival with model simplification: model selection table (for model specifications, see code)

```{r coxme-F1-females-B}

# coxF1.f1 without 3-way interaction par_eff:par_temp:par_diet
coxF1.f2 <- coxme(Surv(lifespan, status) ~ par_eff*par_temp*par_diet - par_eff:par_temp:par_diet + (1|specificvial), 
  data = subset(survivalF1, sex == "f"))

# Model summary (custom function extracts 3 digit z- and p-value)
extract_coxme_table <- function (mod){
        beta <- mod$coefficients #$fixed is not needed
        nvar <- length(beta)
        nfrail <- nrow(mod$var) - nvar
        se <- sqrt(diag(mod$var)[nfrail + 1:nvar])
        z<- round(beta/se, 3)
        p<- signif(1 - pchisq((beta/se)^2, 1), 3)
        table=data.frame(cbind(z,p))
        return(table)
    }
#extract_coxme_table(coxF1.f2)
#summary(coxF1.f2, digits = 3)

#- par_temp:par_diet
coxF1.f3 <- coxme(Surv(lifespan, status) ~ par_eff*par_temp*par_diet - par_eff:par_temp:par_diet - par_temp:par_diet + (1|specificvial), 
  data = subset(survivalF1, sex == "f"))
#summary(coxF1.f3, digits = 3)

#- par_temp:par_diet 
coxF1.f4 <- coxme(Surv(lifespan, status) ~ par_eff*par_temp*par_diet - par_eff:par_temp:par_diet - par_temp:par_diet - par_eff:par_diet + (1|specificvial), 
  data = subset(survivalF1, sex == "f"))
#summary(coxF1.f4, digits = 3)

#- par_eff:par_temp (no interactions left)
coxF1.f5 <- coxme(Surv(lifespan, status) ~ par_eff+par_temp+par_diet + (1|specificvial), 
  data = subset(survivalF1, sex == "f"))
#summary(coxF1.f5, digits = 3)

#- par_diet 
coxF1.f6 <- coxme(Surv(lifespan, status) ~ par_eff+par_temp + (1|specificvial), 
  data = subset(survivalF1, sex == "f"))
#summary(coxF1.f6, digits = 3)

#- par_temp
coxF1.f7 <- coxme(Surv(lifespan, status) ~ par_eff + (1|specificvial), 
  data = subset(survivalF1, sex == "f"))
#summary(coxF1.f7, digits = 3)

# only with par_diet
coxF1.f7b <- coxme(Surv(lifespan, status) ~ par_diet + (1|specificvial), 
  data = subset(survivalF1, sex == "f"))
#summary(coxF1.f7b, digits = 3)

# NULL
coxF1.f8 <- coxme(Surv(lifespan, status) ~ 1 + (1|specificvial), 
  data = subset(survivalF1, sex == "f"))
#summary(coxF1.f8, digits = 3)

# Model comparison
AIC_addCOX_femaleF1 <- model.sel(coxF1.f1, coxF1.f2, coxF1.f3, coxF1.f4, coxF1.f5, coxF1.f6, coxF1.f7, coxF1.f8, rank = "AIC")
AIC_addCOX_female_resF1 <- as.data.frame(AIC_addCOX_femaleF1)[9:12]
AIC_addCOX_female_resF1
```


### 2) F1 males, Cox PH: Model summary (Table S7) and proportionality assumption test

```{r coxme-F1-males}

# Model specification
coxF1.m <- coxme(Surv(lifespan, status) ~ par_eff*par_temp*par_diet + (1|specificvial), 
  data= subset(survivalF1, sex== "m")) 

# Model summary
#extract_coxme_table(coxF1.m)
summary(coxF1.m, digits = 3)

cox.zph(coxph(Surv(lifespan, status) ~ par_eff*par_temp*par_diet, data = subset(survivalF1, sex== "m")))
```

### Male F1 survival with model simplification: model selection table (for model specifications, see code)

```{r coxme-F1-males-B}

# coxF1.m without 3-way interaction par_eff:par_temp:par_diet
coxF1.m2 <- coxme(Surv(lifespan, status) ~ par_eff*par_temp*par_diet - par_eff:par_temp:par_diet + (1|specificvial), 
  data = subset(survivalF1, sex == "m"))

# Model summary (custom function extracts 3 digit z- and p-value)
extract_coxme_table <- function (mod){
        beta <- mod$coefficients #$fixed is not needed
        nvar <- length(beta)
        nfrail <- nrow(mod$var) - nvar
        se <- sqrt(diag(mod$var)[nfrail + 1:nvar])
        z<- round(beta/se, 3)
        p<- signif(1 - pchisq((beta/se)^2, 1), 3)
        table=data.frame(cbind(z,p))
        return(table)
    }
#extract_coxme_table(coxF1.m2)
#summary(coxF1.m2, digits = 3)

#- par_temp:par_diet
coxF1.m3 <- coxme(Surv(lifespan, status) ~ par_eff*par_temp*par_diet - par_eff:par_temp:par_diet - par_temp:par_diet + (1|specificvial), 
  data = subset(survivalF1, sex == "m"))
#summary(coxF1.m3, digits = 3)

#- par_temp:par_diet 
coxF1.m4 <- coxme(Surv(lifespan, status) ~ par_eff*par_temp*par_diet - par_eff:par_temp:par_diet - par_temp:par_diet - par_eff:par_diet + (1|specificvial), 
  data = subset(survivalF1, sex == "m"))
#summary(coxF1.m4, digits = 3)

#- par_eff:par_temp (no interactions left)
coxF1.m5 <- coxme(Surv(lifespan, status) ~ par_eff+par_temp+par_diet + (1|specificvial), 
  data = subset(survivalF1, sex == "m"))
#summary(coxF1.m5, digits = 3)

#- par_diet 
coxF1.m6 <- coxme(Surv(lifespan, status) ~ par_eff+par_temp + (1|specificvial), 
  data = subset(survivalF1, sex == "m"))
#summary(coxF1.m6, digits = 3)

#- par_temp
coxF1.m7 <- coxme(Surv(lifespan, status) ~ par_eff + (1|specificvial), 
  data = subset(survivalF1, sex == "m"))
#summary(coxF1.m7, digits = 3)

# NULL
coxF1.m8 <- coxme(Surv(lifespan, status) ~ 1 + (1|specificvial), 
  data = subset(survivalF1, sex == "m"))
#summary(coxF1.m8, digits = 3)

# Model comparison
AIC_addCOX_maleF1 <- model.sel(coxF1.m, coxF1.m2, coxF1.m3, coxF1.m4, coxF1.m5, coxF1.m6, coxF1.m7, coxF1.m8, rank = "AIC")
AIC_addCOX_male_resF1 <- as.data.frame(AIC_addCOX_maleF1)[9:12]
AIC_addCOX_male_resF1

```
